An AI Success Story in Metaphysics: \\ the Case of the Inconsistency in G\"odel's Ontological Argument

### 2016-01-11 Papers/InProgress/IJCAI/paper.tex

\typeout{Towards Artificial Intelligence for Metaphysics: \\ the Case of the Inconsistency in G\"odel's Ontological Argument}

@@

\title{Towards Artificial Intelligence for Metaphysics: \\ the Case of the Inconsistency in G\"odel's Ontological Argument}
\author{Christoph Benzm\"uller\thanks{German Research Foundation DFG \ldots} and Bruno Woltzenlogel Paleo}
\author{}

@@

On the technical side, the quest for constructing a compelling refutation, 
capable of convincing also human non-experts, led us to improve the 
syntax of the embedding of modal logics in
Isabelle/HOL (as discussed in Section  \ref{ToDo}). With the new syntax, a nearly perfect match
between the original pen and paper presentations and our encoding in
Isabelle/HOl is feasible. A more user-friendly syntax, as reported here, is clearly an important prerequisite
for promoting the theorem proving technology employed here to a wider community of
philosophers, who are not necessarily experts in automated reasoning or higher-order logic.

Another novel contribution reported here (in Section \ref{ToDo}) is the implementation of an alternative embedding for the only slightly stronger modal logic \SFiveU. Our experiments have shown that the new embedding is more efficient, as the following two open problems can now be solved:
\begin{itemize}
\item Proving the final theorem T3 \textit{``Necessarily, there
    exists God''} fully automatically and directly from the axioms alone, without relying on the argument's lemmas.
\item Verifying automatically the proof of the modal
  collapse \cite{Sobel}, which is one of the most strongly criticized
  `side-effects' of the argument's axioms.
\end{itemize}

@@

First successful applications of theorem proving technology in
metaphysics were reported by Oppenheimer and
Zalta~\shortcite{oppenheimera11}, who coined the term \textit{Computational Metaphysics} for this new research area and employed the first-order
\textsc{Prover9} \cite{Prover9} in their experiments. Later on, Rushby~\shortcite{rushby13} used the proof assistant \textsc{PVS} \cite{PVS}. Common to both
works is a significant amount of proof-hand-coding work as well as their
focus on a non-modal formalization of St. Anselm's~\shortcite{Anselm} simpler 
and older ontological argument. 
None of these previous works formalises and automates variants of \emph{higher-order} and \emph{modal} logics, which are, however, crucial
for G\"{o}del's more complex ontological argument.

@@

St. Anselm's ontological argument \cite{Proslogion} can be regarded as the ancestor of modern ontological arguments such as G\"odel's. In the millenium between Anselm and G\"odel, many philosophers modified and arguably improved Anselm's argument. Of particular importance to G\"odel was the work of Leibniz \cite{ToDo}. 
G\"odel's manuscript (Figure \ref{GoedelScript}) can be considered a translation of Leibniz's presentation of the argument into modern modal logic. G\"odel shared his manuscript with Scott, who shared a slightly different version with a larger public. Scott's version of the axioms and definitions, formalized in Isabelle, is shown in Figure \ref{ScottS5}. The main difference to G\"odel's version is the addition of a conjunct in the definition of \emph{essence}. G\"odel's different definition of essence can be seen either in his manuscript (Figure \ref{GoedelScript}) or, in more modern notation, in the Isabelle formalization shown in Figure \ref{nconsistencyIsabelleK}. For Scott, an essential property of an individual must be possessed by him/her. For G\"odel, this is not required. This difference has been considered inessential and merely an oversight by many. For instance, Hazen 1998, p.365 \cite[p.365]{Hazen1998} states that ``G\"odel left this clause out in [11], but this appears to have been an oversight--it is included in related manuscripts''. However, the omitted conjunct is in fact crucial. Without it, G\"odel's original axioms are inconsistent. With it, Scott's axioms are consistent\footnote{In personal communication, Dana Scott confirmed that he was unaware G\"odel's axioms were inconsistent.}.

@@

Logic textbooks \cite{ToDo: which} commonly utilize classical higher-order logic (HOL)
\cite{Church} as a meta-language to introduce the syntax and the
semantics of object logics of interest, in which reasoning
problems in concrete application domains can be modeled and solved
with pen and paper. In fact, this approach can also be followed for even very challenging
object logics to
enable interactive and automated theorem proving with existing theorem provers for classical
higher-order logic.

@@

For a computational analysis of G\"odel's ontological argument in this
approach the embedding of higher-order modal logics (HOML) such as K,
KB and S5 with various domain conditions (possibilist and actualist)
is required. This idea has been successfully followed in related work
\cite{C40}. The embedding of higher-order modal logic is in fact
straightforward. Formulas in HMOL are \emph{lifted}, translated to predicates
over worlds, which are themselves explicitly represented as
terms. The logical constants of HMOL are translated to HOL terms in such a way that, for instance, 
%$\neg \varphi$, $\varphi\vee\psi$
$\Box \varphi$ and $\Diamond \varphi$ are mapped, respectively, to the HOL formulas $\forall w. (r w_0 w) \imp (\varphi w)$ and $\exists w. (r w_0 w) \wedge (\varphi w)$. This form of embedding is precisely the well-known standard translation \cite{Ohlbach,ModalLogicPatrickBlackburn},
which is here (intra-logically) realized in HOL by stating a set of
equations defining the logical constants (Fig. ~\ref{QMLS5}). The resulting object logic is the HOML K with rigid terms and constant domains (possibilist quantifiers). Other logics (e.g. KB, S5) can be embedded by adding axioms that restrict the accessibility relation $r$. Varying domains and actualist quantifiers can be simulated by means of an existence predicate, which can be used to guard the quantifiers. Therefore, the embedding approach is very flexible.

@@

The improved embedding of \SFiveU in Isabelle is shown in Fig.~\ref{QMLS5}.
\begin{figure}
\centerline{\includegraphics[width=\columnwidth]{./Images/QMLS5.png}}
\caption{Improved Embedding of \SFiveU} \label{QMLS5}
\end{figure}

@@

With the improved embedding for \SFiveU, the final theorem T3 (\textit{``Necessarily, there
exists God''}) can be derived from Scott's version of the axioms fully automatically. 
The fully automatic proof has been generated by the theorem prover \textsc{Leo-II}~\cite{C26} 
and subsequently verified in the proof assistant Isabelle/HOL~\cite{NPW02}, as shown in Fig.~\ref{ScottS5}. 
On the other hand, with the embedding used in \shortcite{C40}, 
the provers had to be given the intermediate theorem T2 and the corollary C in order to manage to prove T3.

Another evidence that the new embedding provides a significant performance boost 
is the successful automatic verification in Isabelle/HOL (with its automatic tactic Meson) of the modal
collapse \cite{Sobel}, which is one of the most strongly criticized
`side-effects' of G\"odel's and Scott's variants of the proof. In previous work \shortcite{C40}
the modal collapse has been proven by the higher-order provers
\textsc{Satallax} \cite{Satallax} and \textsc{Leo-II}, but a fully automatic
verification in the highly trusted Isabelle/HOL still failed \cite{ArchiveFormalProofsGoedelGod}. The success with the new embedding can be seen in Fig.~\ref{ScottS5}.

\begin{figure}
\centerline{\includegraphics[width=\columnwidth]{./Images/ScottS5.png}}
\caption{Full Automation of T3 and MC in \SFiveU } \label{ScottS5}
\end{figure}
\marginpar{ToDo: this picture has to be replaced by one without Isabelle errors}

@@

In the embedding presented here, in Fig.~\ref{QMLS5}, advanced syntax-sugaring effects provided by Isabelle were used to prevent the issues enumerated above. The possibility to define boldface connectives allow us to drop the prefix; ``\texttt{binder}'' annotations enable modal quantifiers to be used in the standard binding way and reduce the need for explicit lambda abstractions; and a careful choice of priorities for infix connectives gives the parenthesis conventions that are more familiar to the user. As desired, the definition of essence in Fig.~\ref{QMLS5} is undeniably more immediately recognizable and comprehensible than the definition in Fig.~\ref{UglyEssence}. The embedding technique is now completely transparent to the user.

@@

In the typical workflow of trying to prove a conjecture with a theorem prover, it is customary to check the consistency of the axioms first. For if the axioms are inconsistent, anything (including the conjecture) would be trivially derivable in classical logic (\emph{ex falso quodlibet}). Surprisingly, when this routine check was performed on G\"odel's axioms \cite{ECAI}, the LEO-II prover claimed that the axioms were inconsistent. Unfortunately, the refutation generated by LEO-II was barely human-readable. The text file was 153 lines\footnote{Long lines with an average of 184 characters per line.} long and machine-oriented proof calculus (higher-order resolution \cite{}) and syntax (TPTP THF \cite{}) were used. Part of the file is displayed in Fig.~\ref{LEO-Proof}.

>>>>

\typeout{An AI Success Story in Metaphysics: \\ the Case of the Inconsistency in G\"odel's Ontological Argument}

@@

\title{An AI Success Story in Metaphysics: \\ the Case of the Inconsistency in G\"odel's Ontological Argument}
\author{Christoph Benzm\"uller\thanks{German Research Foundation DFG \ldots} and Bruno Woltzenlogel Paleo}
\author{}

@@

On the technical side, the quest for constructing a compelling refutation, 
capable of convincing also human non-experts, led us to improve the 
syntax of the embedding of modal logics in
Isabelle/HOL (as discussed in Section  \ref{ToDo}). With the new
syntax, a (nearly) perfect match
between the original pen and paper presentations and our encoding in
Isabelle/HOl is feasible. A more user-friendly syntax, as reported here, is clearly an important prerequisite
for promoting the theorem proving technology employed here to a wider community of
philosophers, who are not necessarily experts in automated reasoning or higher-order logic.

Another novel contribution reported here (in Section \ref{ToDo}) is
the implementation of an alternative embedding for the only slightly
stronger modal logic \SFiveU. Our experiments have shown that the new
embedding is more efficient, as the following two, previously open problems can now be solved:
\begin{itemize}
\item Proving the final theorem T3 \textit{``Necessarily, there
    exists God''} fully automatically and directly from the axioms
  alone, without relying on the argument's lemmas (i.e.~without
  relying on intermediate argumentation steps).
\item Verifying automatically (in Isabelle/HOL) the proof of the modal
  collapse \cite{Sobel}, which is one of the most strongly criticized
  `side-effects' of the argument's axioms.
\end{itemize}

@@

First successful applications of theorem proving technology in
metaphysics were reported by Oppenheimer and
Zalta~\shortcite{oppenheimer11}, who coined the term \textit{Computational Metaphysics} for this new research area and employed the first-order
\textsc{Prover9} \cite{prover9-mace4} in their experiments. Later on, Rushby~\shortcite{rushby13} used the proof assistant \textsc{PVS} \cite{cade92-pvs}. Common to both
works is a significant amount of proof-hand-coding work as well as their
focus on a non-modal formalization of St. Anselm's~\shortcite{Proslogion} simpler 
and older ontological argument. 
None of these previous works formalises and automates variants of \emph{higher-order} and \emph{modal} logics, which are, however, crucial
for G\"{o}del's more complex ontological argument.

@@

St. Anselm's ontological argument \cite{Proslogion} can be regarded as the ancestor of modern ontological arguments such as G\"odel's. In the millenium between Anselm and G\"odel, many philosophers modified and arguably improved Anselm's argument. Of particular importance to G\"odel was the work of Leibniz \cite{ToDo}. 
G\"odel's manuscript (Figure \ref{GoedelScript}) can be considered a translation of Leibniz's presentation of the argument into modern modal logic. G\"odel shared his manuscript with Scott, who shared a slightly different version with a larger public. Scott's version of the axioms and definitions, formalized in Isabelle, is shown in Figure \ref{Scott_S5}. The main difference to G\"odel's version is the addition of a conjunct in the definition of \emph{essence}. G\"odel's different definition of essence can be seen either in his manuscript (Figure \ref{GoedelScript}) or, in more modern notation, in the Isabelle formalization shown in Figure \ref{nconsistencyIsabelleK}. For Scott, an essential property of an individual must be possessed by him/her. For G\"odel, this is not required. This difference has been considered inessential and merely an oversight by many. For instance, Hazen 1998, p.365 \cite[p.365]{Hazen1998} states that ``G\"odel left this clause out in [11], but this appears to have been an oversight--it is included in related manuscripts''. However, the omitted conjunct is in fact crucial. Without it, G\"odel's original axioms are inconsistent. With it, Scott's axioms are consistent\footnote{In personal communication, Dana Scott confirmed that he was unaware G\"odel's axioms were inconsistent.}.

@@

Logic textbooks \cite{ToDo:which} commonly utilize classical higher-order logic (HOL)
\cite{Church} as a meta-language to introduce the syntax and the
semantics of object logics of interest, in which reasoning
problems in concrete application domains can be modeled and solved
with pen and paper. In fact, this approach can also be followed for even very challenging
object logics to
enable interactive and automated theorem proving with existing theorem provers for classical
higher-order logic.

@@

For a computational analysis of G\"odel's ontological argument in this
approach the embedding of higher-order modal logics (HOML) such as K,
KB and S5 with various domain conditions (possibilist and actualist)
is required. This idea has been successfully followed in related work
\cite{C40}. The embedding of higher-order modal logic is in fact
straightforward. Formulas in HMOL are \emph{lifted}, translated to predicates
over worlds, which are themselves explicitly represented as
terms. The logical constants of HMOL are translated to HOL terms in such a way that, for instance, 
%$\neg \varphi$, $\varphi\vee\psi$
$\Box \varphi$ and $\Diamond \varphi$ are mapped, respectively, to the HOL formulas $\forall w. (r w_0 w) \imp (\varphi w)$ and $\exists w. (r w_0 w) \wedge (\varphi w)$. This form of embedding is precisely the well-known standard translation \cite{Ohlbach,ModalLogicPatrickBlackburn},
which is here (intra-logically) realized in HOL by stating a set of
equations defining the logical constants (Fig. ~\ref{QML_S5}). The resulting object logic is the HOML K with rigid terms and constant domains (possibilist quantifiers). Other logics (e.g. KB, S5) can be embedded by adding axioms that restrict the accessibility relation $r$. Varying domains and actualist quantifiers can be simulated by means of an existence predicate, which can be used to guard the quantifiers. Therefore, the embedding approach is very flexible.

@@

The improved embedding of \SFiveU in Isabelle is shown in Fig.~\ref{QML_S5}.
\begin{figure}
\centerline{\includegraphics[width=\columnwidth]{./Images/QML_S5.png}}
\caption{Improved Embedding of \SFiveU} \label{QML_S5}
\end{figure}

@@

With the improved embedding for \SFiveU, the final theorem T3 (\textit{``Necessarily, there
exists God''}) can be derived from Scott's version of the axioms fully automatically. 
The fully automatic proof has been generated by the theorem prover \textsc{Leo-II}~\cite{C26} 
and subsequently verified in the proof assistant Isabelle/HOL~\cite{NPW02}, as shown in Fig.~\ref{Scott_S5}. 
On the other hand, with the embedding used in \shortcite{C40}, 
the provers had to be given the intermediate theorem T2 and the corollary C in order to manage to prove T3.

Another evidence that the new embedding provides a significant performance boost 
is the successful automatic verification in Isabelle/HOL (with its automatic tactic Meson) of the modal
collapse \cite{Sobel}, which is one of the most strongly criticized
`side-effects' of G\"odel's and Scott's variants of the proof. In previous work \shortcite{C40}
the modal collapse has been proven by the higher-order provers
\textsc{Satallax} \cite{Satallax} and \textsc{Leo-II}, but a fully automatic
verification in the highly trusted Isabelle/HOL still failed \cite{ArchiveFormalProofsGoedelGod}. The success with the new embedding can be seen in Fig.~\ref{Scott_S5}.

\begin{figure}
\centerline{\includegraphics[width=\columnwidth]{./Images/Scott_S5.png}}
\caption{Full Automation of T3 in \SFiveU; Verification of MC} \label{Scott_S5}
\end{figure}
\marginpar{ToDo: this picture has to be replaced by one without Isabelle errors}

@@

In the embedding presented here, in Fig.~\ref{QML_S5}, advanced syntax-sugaring effects provided by Isabelle were used to prevent the issues enumerated above. The possibility to define boldface connectives allow us to drop the prefix; ``\texttt{binder}'' annotations enable modal quantifiers to be used in the standard binding way and reduce the need for explicit lambda abstractions; and a careful choice of priorities for infix connectives gives the parenthesis conventions that are more familiar to the user. As desired, the definition of essence in Fig.~\ref{QML_S5} is undeniably more immediately recognizable and comprehensible than the definition in Fig.~\ref{UglyEssence}. The embedding technique is now completely transparent to the user.

@@

In the typical workflow of trying to prove a conjecture with a theorem prover, it is customary to check the consistency of the axioms first. For if the axioms are inconsistent, anything (including the conjecture) would be trivially derivable in classical logic (\emph{ex falso quodlibet}). Surprisingly, when this routine check was performed on G\"odel's axioms \cite{C40}, the LEO-II prover claimed that the axioms were inconsistent. Unfortunately, the refutation generated by LEO-II was barely human-readable. The text file was 153 lines\footnote{Long lines with an average of 184 characters per line.} long and machine-oriented proof calculus (higher-order resolution \cite{}) and syntax (TPTP THF \cite{}) were used. Part of the file is displayed in Fig.~\ref{LEO-Proof}.
